---
title: "Understanding sources of uncertainty in Ecological Niche Models"
author: "Sara Mortara"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    theme: cosmo
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r pkg}
library(ggplot2)
library(dplyr)
library(data.table)
library(stringr)
library(wesanderson)
library(knitr)
library(raster)
library(rasterVis)
library(sp)
library(ggcorrplot)
## for model selection
library(lme4)
library(bbmle)
library(MuMIn)
# functions
source("R/sun_plot.R")
source("R/read_sun.R")
```



```{r prep}
# species data
sp.trat <- list.files("results/uncertainty")
sp.df <- read.csv("data/epiphyte_species.csv", as.is = TRUE)
fam <- sort(table(sp.df$family), decreasing = TRUE)
fam.n <- as.numeric(fam)
# color palette
n.col <- 4
cores <-  wes_palette("Zissou1", n.col,
                      type = "continuous")
# algorithms
nome.algo <- c("Bioclim", "GLM", "Maxent", "Random Forest", "SVM")

# final stats
stats_rm <- read.csv("results/final_tables/stats_row_mean.csv")
stats_rmc <- read.csv("results/final_tables/stats_row_mean_cut.csv")
stats_rm$size <- as.factor(stats_rm$size)
stats_rmc$size <- as.factor(stats_rmc$size)

# occs
# species names
spp_names <- list.files("./data/epiphyte_occ/")
# listing .csv files
occ_files <- list()
for (i in 1:length(spp_names)) {
  occ_files[[i]] <- list.files(path = paste0("./data/epiphyte_occ/", spp_names[i]),
                               pattern = '.*0.csv',
                               full.names = TRUE)
}
# reading csv files
occ_data <- lapply(unlist(occ_files),
                   fread, sep = ',')

```


# Goal

Our main goal is to quantify uncertainty in ecological niche models (ENMs). First, we will explore uncertainty in ecological niche models of epiphyte species from the Atlantic Forest using different algorithms and experimental scenarios (sample size, spatial autocorrelation and dimensionality). 

# Methods

## Biotic variables

We are using occurrence data from epiphyte species from the Atlantic Forest published in the database by Ramos et al. (2019) *Ecology*. We selected only Angiosperm epiphytes endemic to the Atlantic Forest (according to Brazilian Flora 2020 -- the official Brazilian plant list). We are using occurrence data from 33 species belonging to Bromeliaceae (N=`r fam.n[1]`), Orchidaceae (N=`r fam.n[2]`), Cactaceae (N=`r fam.n[3]`), Gesneriaceae (N=`r fam.n[4]`) Melastomataceae (N=`r fam.n[5]`) and Piperaceae (N=`r fam.n[6]`) families. 

```{r tab.sp}
sp.list <- sp.df$sp
sp.df$sp <- gsub("_", " ", sp.df$sp)
sp.df$sp <- paste0("*", sp.df$sp, "*")
kable(sp.df,
      col.names = c("Family", "Species"),
      caption = "Table 1. List of species included in this report.")
```


## Abiotic variables

We are using environmental variables from World Clim at 2.5 ° for South America. We selected all variables except: Mean Temperature of Wettest Quarter (BIO08), Mean Temperature of Driest Quarter (BIO09), Precipitation of Warmest Quarter (BIO18), Precipitation of Coldest Quarter (BIO19). We performed a PCA with all environmental variables to build two environmental scenarios: low and high dimensionality. Low dimensionality scenario is represented by the first four PCA axes (> 90% of variance), whereas the high dimensionality scenario is represented by the eight fisrt axes (> 98% of variance).

## Experimental design

From 33 species we generated different experimental scenarios with different sample sizes (10, 20, 50 and 100% of all occurrence data points) and different patterns of spatial aggregation (clump and no clump). For each treatment N=`r length(sp.trat)`  we modeled ecological niche using the following algorithms: `r nome.algo`. So far, we only run models for the low dimensionality scenario. We separed data into four partitions and generated 1,000 pseudo absences around a buffer based on mean distance from points, cutting with a minimum distance for the exclusion buffer. 

## Statistical analysis

In order to understand how the proposed metrics reflect model uncertainty, we fitted a generalized linear mixed model (GLMM) with a binomial response error to model the variation in uncertainty metrics (M1 and M2). We included clumping, sample size and algorithm as fixed effects and species as a random effect. We selected the best model by using using model and Akaike Information Criterion (AIC) to make statistical inference.In addition, we used partial R^${2}$ (sensu Nakagawa) to quantify relative contribution of fixed and random effects. 

# Results

Legends

- Clumping:
  - C = clump
  - N = no clump

- Sample size:
  - 010 = 10% sample size
  - 020 = 20% sample size
  - 050 = 50% sample size
  - 100 = 100% sample size

- Uncertainty metrics
  - M1 = N of good points/N total
  - M2 = N of good points/N bad points
  - M3 = N of good points/(N bad points + N of good points) 
  

## Understanding model metrics

### AUC values for each treatment

```{r auc}
ggplot(stats_rm, aes(x = algorithm, y = AUC, fill = size)) +
  geom_boxplot() +
  facet_grid(clump ~ .) +
  scale_fill_manual(values = cores) +
  theme_classic()
```


### TSS values for each treatment

```{r tss}
ggplot(stats_rm, aes(x = algorithm, y = TSS, fill = size)) +
  geom_boxplot() +
  facet_grid(clump ~ .) +
  scale_fill_manual(values = cores) +
  theme_classic()
```

### pROC values for each treatment

All pROC values are significant. We should check this. 

```{r proc}
ggplot(stats_rm, aes(x = algorithm, y = pROC, fill = size)) +
  geom_boxplot() +
  facet_grid(clump ~ .) +
  scale_fill_manual(values = cores) +
  theme_classic()
```


### Omission values for each treatment

```{r omission}
ggplot(stats_rm, aes(x = algorithm, y = omission, fill = size)) +
  geom_boxplot() +
  facet_grid(clump ~ .) +
  scale_fill_manual(values = cores) +
  theme_classic()
```

## Understanding overall uncertainty

### M1 = good/total

```{r m1}
ggplot(stats_rm, aes(x = algorithm, y = M1, fill = size)) +
  geom_boxplot() +
  facet_grid(clump ~ .) +
  scale_fill_manual(values = cores) +
  theme_classic()
```

### M2 = good/bad

```{r m2}
ggplot(stats_rm, aes(x = algorithm, y = M2, fill = size)) +
  geom_boxplot() +
  labs(y = "M2 (log)") +
  facet_grid(clump ~ .) +
  scale_fill_manual(values = cores) +
  scale_y_log10() +
  theme_classic()
```

### M3 = good/(good + bad)

```{r m3}
ggplot(stats_rm, aes(x = algorithm, y = M3, fill = size)) +
  geom_boxplot() +
  facet_grid(clump ~ .) +
  scale_fill_manual(values = cores) +
  theme_classic()
```

## Correlation among all metrics

```{r cor}
metrics <- stats_rm[,c('AUC', 'TSS', 'pROC', 'M1', 'M3', 'omission')]
correlation <- cor(metrics, use = 'complete.obs')

ggcorrplot(correlation, 
           hc.order = TRUE, 
           type = "lower",
           lab = TRUE)
  
```

## Variance partitioning of uncertainty metrics with GLMMs

```{r mod.M1}
stats_rm$sp.gen <- paste(stats_rm$genus, stats_rm$epithet, sep = "_")
# f <- M1 ~ clump*size*algorithm + (1|sp.gen)
# m <- buildmer(f,data=stats_rm,direction='order',control=lmerControl(optimizer='bobyqa'))

m01 <- lmer(M1 ~ clump + size + algorithm + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m02 <- lmer(M1 ~ clump + size + (1|sp.gen),
             data = stats_rm, family = "binomial")
m03 <- lmer(M1 ~ clump + algorithm + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m04 <- lmer(M1 ~ size + algorithm + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m05 <- lmer(M1 ~ size + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m06 <- lmer(M1 ~ algorithm + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m07 <- lmer(M1 ~ clump + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m00 <- lmer(M1 ~ 1 + (1|sp.gen), 
             data = stats_rm, family = "binomial") 
AICtab(m01, m02, m03, m04, m05, m06, m07, m00, 
       base = TRUE, weights = TRUE)


```



```{r mod.M3}
m01b <- lmer(M3 ~ clump + size + algorithm + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m02b <- lmer(M3 ~ clump + size + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m03b <- lmer(M3 ~ clump + algorithm + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m04b <- lmer(M3 ~ size + algorithm + (1|sp.gen),
             data = stats_rm, family = "binomial")
m05b <- lmer(M3 ~ size + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m06b <- lmer(M3 ~ algorithm + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m07b <- lmer(M3 ~ clump + (1|sp.gen), 
             data = stats_rm, family = "binomial")
m00b <- lmer(M3 ~ 1 + (1|sp.gen), 
             data = stats_rm, family = "binomial") 
AICtab(m01b, m02b, m03b, m04b, m05b, m06b, m07b, m00b, 
       base = TRUE, weights = TRUE)
```

## Suitability vs. uncertainty plots

### Looking at C 100 species

#### Bioclim C_100

```{r bio100, fig.width=3, fig.height=2.5}
bio_100 <- read_sun(algo = "bioclim", 
                    cod = "C_100", 
                    type = "raw_mean")

for (i in 1:length(bio_100)) {
sun_plot(bio_100[[i]], 
         title = paste(sp.list))
}
```


#### GLM C_100

```{r glm100, fig.width=3, fig.height=2.5}
glm_100 <- read_sun(algo = "glm", 
                    cod = "C_100", 
                    type = "raw_mean")

for (i in 1:length(glm_100)) {
sun_plot(glm_100[[i]], 
         title = paste(sp.list))
}
```

#### Maxent C_100

```{r max100, fig.width=3, fig.height=2.5}
max_100 <- read_sun(algo = "maxent", 
                    cod = "C_100", 
                    type = "raw_mean")

for (i in 1:length(max_100)) {
sun_plot(max_100[[i]], 
         title = paste(sp.list))
}
```

#### Random Forest C_100

```{r rf100, fig.width=3, fig.height=2.5}
rf_100 <- read_sun(algo = "rf", 
                    cod = "C_100", 
                    type = "raw_mean")

for (i in 1:length(rf_100)) {
sun_plot(rf_100[[i]], 
         title = paste(sp.list))
}
```

#### Support Vector Maxines C_100

```{r svm100, fig.width=3, fig.height=2.5}
svm_100 <- read_sun(algo = "svmk", 
                    cod = "C_100", 
                    type = "raw_mean")

for (i in 1:length(svm_100)) {
sun_plot(svm_100[[i]], 
         title = paste(sp.list))
}
```


```{r map1}
algo <- "bioclim"
dir <- "models_low"
sp <- sp.trat[1]
occ <- occ_data[[1]]

sun_map(algo = algo, 
        dir = dir, 
        sp = sp,
        occ = occ)
```

